---
title: "assignment_1"
author: "lloyd hill"
output:
  pdf_document: default
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
tgpp <- read.csv("https://raw.githubusercontent.com/dmcglinn/quant_methods/gh-pages/data/tgpp.csv")

```

1. What are the names of the columns in this dataset?

```{r}
#I used the names() function to call the column headings from the dataset. 

names(tgpp)

```

2. How many rows and columns does this data file have?

```{r}
#Here, I used the dimensions function, dim(), to count the number of rows and columns in the dataset. There are 4,080 rows and 11 columns. 

dim(tgpp)
```

3. What kind of object is each data column? Hint: checkout the function sapply().

```{r}
#With the sapply() function, I selected the data frame of interest, tgpp, and the 'class' function. This provides the data types of each column. This dataset contains two types of data -- integer and numeric.  

sapply(tgpp, class)
```

4. What are the values of the the datafile for rows 1, 5, and 8 at columns 3, 7, and 10

```{r}
#I used the square brackets to select the desired values from the dataset. Using the c() function I was able to select multiple rows and columns. Here I've selected the 1st, 5th, and 8th rows from columns 3, 7, and 10. 

tgpp[c(1,5,8),c(3,7,10)]
```

5. Create a pdf of the relationship between the variables "scale" and "richness". Scale is the area in square meters of the quadrat in which richness was recorded. Be sure to label your axes clearly, and choose a color you find pleasing for the points. To get a list of available stock colors use the function colors(). 

```{r}
pdf(file="../tgpp_plot.pdf")
plot(tgpp$scale,tgpp$richness, col = "blue", ylab = "Species Richness", xlab = "Quadrat Area", log =  "xy", main = "Tallgrass Prairie Plant Richness")
dev.off()
```

