---
title: "R intermediate"
author: "lloyd hill"
date: "1/24/2019"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

> Premature optimization is the root of all evil -- Donald Knuth

The humble for loop is often considered distasteful by seasoned programmers because it is inefficient; however, the for loop is one of the most useful and generalizable programming structures in R. If you can learn how to construct and understand for loops then you can code almost any iterative task. Once your loop works you can always work to optimize your code and increase its efficiency.

Before attempting these exercises you should review the lesson[R intermediate](../lessons/R_intermediate) in which loops were covered. 

Examine the following for loop, and then complete the exercises

```{r}
data(iris)
head(iris)

sp_ids = unique(iris$Species)

output = matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(output) = sp_ids
colnames(output) = names(iris[ , -ncol(iris)])

for(i in seq_along(sp_ids)) {
    iris_sp = subset(iris, subset=Species == sp_ids[i], select=-Species)
    for(j in 1:(ncol(iris_sp))) {
        x = 0
        y = 0
        if (nrow(iris_sp) > 0) {
            
            for(k in 1:nrow(iris_sp)) {
              
                x = x + iris_sp[k, j]
                y = y + 1
            }
            output[i, j] = x / y 
        }
    }
}

output

```
##Excercises
###Iris loops

1. Describe the values stored in the object `output`. In other words what did the loops create?

```{r}

#The above function creates a new matrix from the iris dataset. It calculated the means of different plant measurements (i.e., sepal length) for each species (i.e., I. setosa). 

```
2. Describe using pseudo-code how `output` was calculated, for example,
```{r}
#Create an empty matrix ('output') with the proper dimensions and headers (rows = 3 species; columns = 4measurement type)

#First loop: for i in 1:3 (number of species), use the subset function to create a subset of the iris data for each species. assign these subseted, species-specific dataframes to 'iris_sp'
  
    #Second loop: for j in 1:4 (number of measurements (eg, sepal length)). Create empty objects 'x' and 'y'.      
    #if() iris_sp  contains more than zero rows, then move to third loop
    
        #Third loop: for k in 1:50 (number of individual observations). 
        # x = x + value at the kth row and jth column.
        # y = y + 1 (calculates the total number of observations, 50)
      
    #Now the 'output' matrix can be populated with the means by dividing x/y from the third loop. These means      
    #are assigned with the 'i' and 'j' indices (output[i,j] = x/y). 

```

3. The variables in the loop were named so as to be vague. How can the objects `output`, `x`, and `y` could be renamed such that it is clearer what is occurring in the loop.

```{r}
#If I were making this function, I would rename 'output' to 'iris_means'; 'x' to 'sum'; and 'y' to 'n'. I believe this would help the reader better follow the code. 
```

4. It is possible to accomplish the same task using fewer lines of code? Please suggest one other way to calculate `output` that decreases the number of loops by 1. 
```{r}
#I removed the last loop which calculated the mean values by dividing the sum of observed values by the number of observations. I then replaced that section with R's built in mean() function.

sp_ids <- unique(iris$Species)

iris_means <- matrix(0, nrow=length(sp_ids), ncol=ncol(iris)-1)
rownames(iris_means) <- sp_ids
colnames(iris_means) <- names(iris[,-ncol(iris)])

for(i in seq_along(sp_ids)) {
    iris_sp <- subset(iris, subset=Species == sp_ids[i], select=-Species)
        for(j in 1:(ncol(iris_sp))) {
            iris_means[i,j] <- mean(iris_sp[,j])
        }
}

print(iris_means)
```


###Sum of a sequence

5. You have a vector `x` with the numbers 1:10. Write a for loop that will produce a vector `y` that contains the sum of `x` up to that index of `x`. So for example the elements of `x` are 1, 2, 3, and so on and the  elements of `y` would be 1, 3, 6, and so on. 

```{r}
x <- 1:10
y <- NULL

for(i in x) {
    y[i] <- sum(x[1:i])
}

print(y)
```
6. Modify your for loop so that if the sum is greater than 10 the value of `y` is set to NA
```{r}
x <- 1:10
y <- NULL

for(i in x) {
    y[i] <- sum(x[1:i])
        if (y[i] > 10) {
            y[i] <- NA
        }
}

print(y)
```

7. Place your for loop into a function that accepts as its argument any vector of arbitrary length and it will return `y`. 
```{r}
#created function called 'my_loop' that creates a vector ('y') which is made up of the sum of all previous values in a sequence of integers. The length of this sequence is defined by 'n'.  

my_loop <- function(n) {
    x <- n[1]:n[length(n)]
    y <- NULL
        for(i in x){
            y[i] <- sum(x[1:i])
        }
    return(y)
    }


A <- c(2, 8, 11)
my_loop(A)
```



